---
title: "Field work Location randomization"
author: "Nelma Péclard"
output: html_document
date: "2024-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Concept

I have 9 locations which need to be sampled at 3 times of day over the course of three months (May, June, July).

Each location will be sampled in the morning, at midday, and in the afternoon.

Each month, each location will be sampled once, at a different time of day each month (e.g. location 1 is sampled in May in the morning, in July at midday, in June in the afternoon).

This means each month, 3 locations will be sampled in the morning, 3 at midday, 3 in the afternoon.

### Code

This code gives me a randomized schedule which states which location will be visited at which time of day each month.

Non-redundancy is ensured using a visit counter matrix.


```{r}
# Define the locations and times of day
locations <- 1:9
times_of_day <- c("morning", "lunch", "afternoon")

# Initialize an empty list to store the randomized schedule
randomized_schedule <- list()

# Initialize a visits counter matrix to keep track of visits per location and time of day
visits_counter <- matrix(0, nrow = length(locations), ncol = length(times_of_day))
colnames(visits_counter) <- times_of_day

# set random seed for reproducibility (obtain same randomized schedule every time)
set.seed(12)

# Loop through each month
for (month in c("May", "June", "July")) {
  # Initialize a data frame to store visits for the current month
  month_schedule <- data.frame(Location = integer(), TimeOfDay = character(), Month = character(), stringsAsFactors = FALSE)
  
  # Loop through each time of day
  for (time in times_of_day) {
    # Sample locations that have not been visited yet at this time of day and in previous months
    available_locations <- locations[visits_counter[,time] == 0] #ensures only locations not yet visited at a specific time of day can be selected
    available_locations <- available_locations[!available_locations %in% month_schedule$Location] #ensures each location is selected only once per month
    sampled_locations <- sample(available_locations, 3, replace = FALSE)
    
    # Update the visits counter for the sampled locations
    visits_counter[sampled_locations, time] <- visits_counter[sampled_locations, time] + 1
    
    # Create a data frame with the sampled locations for the current time of day and month
    time_schedule <- data.frame(Location = sampled_locations, TimeOfDay = rep(time, 3), Month = rep(month, 3), stringsAsFactors = FALSE)
    
    # Append the time_schedule to the month_schedule
    month_schedule <- rbind(month_schedule, time_schedule)
  }
  
  # Append the current month's schedule to the list
  randomized_schedule[[month]] <- month_schedule
}

# Combine the randomized schedules for all months into a single data frame
randomized_schedule <- do.call(rbind, randomized_schedule)

# Display the randomized schedule
print(randomized_schedule)
```
